---
title: 'Class 12 pt.2 HW: Q13, Q14'
author: 'Sion Kang (PID: A17052234)'
date: "2025-02-13"
output: pdf_document
---

## Section 4: Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3 expression.

### Q13
How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

The sample size for each genotype
```{r}
table(expr$geno)
```

**A/A: 108, A/G: 233, G/G: 121**

Finding the median expression levels for each genotype

```{r}
summary(expr$exp[expr$geno == "A/A"])
summary(expr$exp[expr$geno == "A/G"])
summary(expr$exp[expr$geno == "G/G"])
```


**A/A median: 31.25; A/G median: 25.065; G/G median: 20.074**


```{r}
library(ggplot2)
```

Q. 14
Let's make a boxplot

```{r}
ggplot(expr) +
  aes(x=geno, y=exp, fill=geno) +
  geom_boxplot(notch=TRUE)
```
**From this boxplot, we can infer that the expression value is correlated to the genotypes (A/A, G/G). Thus, we can infer that the SNP effects the expression of ORMDL3.**
